Closing the gap between statistical and scientific workflows for improved inference in ecology

### 2025-07-22 revision/response_r1.tex

\begin{mybox}
\emph{L36: These workflow citations are all unpublished (arXiv). I’m fine with that, but consider citing some from other fields that have gone through peer review. The caption of Figure 2 includes more, suggest putting them here too.}
\end{mybox}

Done.

\begin{mybox}
\emph{L103: What does “change the model structure” mean here? Adding constraints and priors? Removing or adding complexity?}
\end{mybox}

...
% Jim: I think the reviewer makes a fair point that this is vague. Can we add something here, or even just refer to section 3.2 where (I think) relevant examples are already given in the final paragraph?

@@

\begin{mybox}
\emph{Fig 2: Caption should mention this is expanded from Fig. 1C.}
\end{mybox}

Done.

\begin{mybox}
\emph{L116, L168: I disagree that “reduce uncertainties” is the goal – instead it is to more accurately quantify it. See L 257 “properly accounting for their uncertainty” for instance.}
\end{mybox}

We have clarified this (\href{file:forecastflows_r1\#lintarget:quant1}{line \ref*{lin:quant1}}, \href{file:forecastflows_r1\#lintarget:quant2}{line \ref*{lin:quant2}}).

@@

\begin{mybox}
\emph{Further, is the proposed workflow to conduct a full-blown simulation calibration study (i.e., simulate-fit-check for capture) many times and across all possible parameter values? Assuming yes, I’d be interested in how something like that would scale for the types of models that are the focus of the paper? Could the authors provide more guidance on how to conduct such a study in this context?}
\end{mybox}

We have clarified the data simulaton step. We believe a first step would be to simulate from one parameter set (\href{file:forecastflows_r1\#lintarget:firstsim}{line \ref*{lin:firstsim}}), and then to repeat the simulations with several parameter sets (\href{file:forecastflows_r1\#lintarget:repeatsim}{line \ref*{lin:repeatsim}}, \href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}).

\begin{mybox}
\emph{4. Gaps in the workflow\\
A big argument in the paper is that the proposed workflow could lessen discrepancies in results between studies of trends (lines 123-128). However, in the workflow there is no mention of transparent communication of study design, which if ignored in analysis, could result in discrepancies in results for studies with similar research questions. Differences in sampling designs must also be considered prior tointegrating different data sources (if integration is even possible). The importance of study design/sampling design and its ties to appropriate modeling is not discussed. Is itimplicitly assumed that there is alignment between design, question, and model in this workflow? Some mention of the importance of study design would strengthen the work.}
\end{mybox}

We have added an entire paragraph to discuss the methodological aspects, including study/sampling design, that should be more transparently shared along with the use of this workflow (\href{file:forecastflows_r1\#lintarget:studydesignandmore}{line \ref*{lin:studydesignandmore}}).

@@

\begin{mybox}
\emph{5. Theoretical examples provided, but no data application of the proposed workflow\\
Sections 3.1 and 3.2 and the Box outline theoretical examples of how the workflow could be applied for trend estimation and forecasting in the context of types of data, but there are no concrete data examples. A comment on how the proposed workflow could be applied to real data and how it would scale with complexity of data and research questions would be a good addition to the paper if it is not feasible to include real-data examples with the general discussion.}
\end{mybox}

% Real-data examples... not do-able!
...

\begin{mybox}
\emph{6. Adopting the workflow\\
Recommendations are made to train ecologists more quantitatively, but there is no suggestion or encouragement of collaboration with statisticians. These types of workflows are generally part of statistical practice, so could an alternative to more quantitative training (which would have to come at the cost of some other aspect of training) be more cross-disciplinary collaboration with statisticians?}
\end{mybox}

We have added thus suggestion (\href{file:forecastflows_r1\#lintarget:collab}{line \ref*{lin:collab}}).

### 2025-07-22 revision/forecastflows_r1.tex
incorporate Jim feedback

\documentclass[11pt]{article}
\usepackage[top=0.70in, bottom=.80in, left=1in, right=1in]{geometry}
\renewcommand{\baselinestretch}{1.1}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{multicol}
\usepackage{tcolorbox}
\usepackage{caption}
\usepackage{todonotes}
\usepackage[font=sf]{caption}
\usepackage{hyperref}
\newcommand{\llabel}[1]{\hypertarget{lintarget:#1}{}\linelabel{lin:#1}}

@@

\llabel{inferentialgoal}The inferential goals of these two modeling paradigms---explanation vs.\ prediction---seem disconnected at first glance, but remains in reality much more blurred. Trend estimation, often with simpler models, tend to focus on an explanatory purpose, but often lacks a clear link with theories that could provide causal interpretation. For forecasting, the use of more complex process-based models for a predictive purpose is often justified by the assumption that ``we must understand to predict", contributing to the confusion between explanation and prediction \citep{Shmueli2010}.

@@

@@

\begin{figure}[h]
	\centering
	\includegraphics{../figures/figure_modeldataspaces_revised}
	\caption{Different trajectories through model-data space during model development, from pre-model/pre-data to post-model/post-data quadrants.
	Panel A serves as a legend, showing where model development should begin (carefully thinking about the hypotheses we want to test) and when inference should be made, for either explanatory purpose (estimation) or forecasting (prediction).
	Panel B illustrates that in current trend estimation approaches, most time is spent in the post-model/post-data quadrant, where models are directly fitted to empirical data---often with no prior evaluation of the model.
	Panel C illustrates the typical process-based model development. A lot of time is spent building and calibrating (fitting) separate submodels (represented by the separated ovals), without clear distinction between model building and model calibration and without evaluating the full model structure a priori. Additionally, the specific hypotheses the model is designed to test are rarely clearly stated.
	Panel D illustrates the workflow highlighted in this paper (see main text and Figure 2 for details). Ideally, it begins in the pre-model/pre-data quadrant, where the research question is defined. The next step moves to the post-model/pre-data quadrant, where a model is built---independently of the data. This is where we should spend a good deal of time evaluating model behavior using simulated data. Once we are satisfied with our model, we can move to the post-model/post-data quadrant, where the model is fitted to real data. Then, we perform retrodictive checks to compare model predictions to observations, which likely give us some feedback to refine our model.}
	\label{fig:modeldata}
\end{figure}

@@

@@

\begin{figure}[h]
	\centering
	\vspace*{-0.1cm}
    \hspace*{-1.5cm}
	\includegraphics{../figures/figure_worflow_wsteps}
	\caption{The workflow we propose here \citep[which builds from recent advances in workflows,][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024} focuses on iterative feedbacks between the research question, model and data. Most research should start at Step 1, with the research question, followed by extensive model building and evaluation using simulated data (Steps 2 and 3) before proceeding to fitting the model on empirical data and examining it through retrodictive checks (Steps 4 and 5). Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.  The integration of observations (the data stream) occurs at Step 4---only after the model has been thoroughly evaluated---and can also highlight opportunities to collect new data and enrich the data stream. This figure is expanded from Figure \ref{fig:modeldata}d.}
	\vspace*{-0.5cm}
	\label{fig:workflow}
\end{figure}

@@

Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to accurately \llabel{quant1}quantify uncertainties and yield new scientific insights. This will help refocus the debate on designing new hypotheses, formulating new questions---and guiding efforts to collect new data. Here, we illustrate how such a workflow could lead to significant improvements in two case studies: (i) estimating global biodiversity trends and (ii) forecasting future species and ecosystem dynamics using process-based models.

@@

Increasing model complexity can be beneficial, especially when it helps better \llabel{quant2}quantify the uncertainty; however, this is not always the outcome. In process-based modeling for climate forecasts, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong disagreements exist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017}.
These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
Some researchers now advocate for simplifying models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. 
If a model becomes too complex, it may become a black box,
and understanding the sources of uncertainty and how they propagate through the model may become nearly impossible (see Box).
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.

@@

Applying the workflow to process-based models would help open the black box.
Each successive step of model development in the workflow may highlight current problems and a path to solutions. Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two, and help expose potential parameter identifiability issues in the model design. 
Uncovering identifiability issues would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity. The workflow may also highlight model problems by requiring more explicit model calibration (i.e. data fitting), which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model. Through the workflow non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity). The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating it through projections.

@@

% Trends - outline current problem
In the current workflow for estimating trends over time, a new model with a new estimate often leads to a paper (see figure A above) because ecologists spend less time interrogating their models with simulated data, or their model performance fit to empirical data. 
The Living Planet Index (LPI, \url{www.livingplanetindex.org}), which aims to include long-term data on vertebrate populations of species across the globe, is emblematic of these conflicting results. % \todo[]{could highlight how influential the LPI has become (recent reports suggest a 73\% ave. pop decline since 1970. This would be massive!}
With updated data released semi-annually alongside new datasets and new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach. For examples using LPI data, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed many previous results \citep{Johnson2024}.

@@

We suggest unified training in our proposed workflow, or a similar one \citep{betanworkflow,Gelman2020, grinsztajn2021,vandeschoot2021}, would focus on learning to generate questions and then models, and then how to simulate data from them. Through this and the use retrodictive checks, most ecologists would be better equipped to think through what parameters are most critical to their question and/or aim (e.g., management), and also gain a much stronger connection to the level of uncertainty in many of ecological estimates. Empiricists would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models. Process-based modelers may start a new generation of simpler models that are more tractable to theoretical ecologists, who may see new bridges from their work to empirical data and forecasting.

\subsection{A tractable alternative to machine learning}

>>>>

\documentclass[11pt]{article}
\usepackage[top=0.70in, bottom=.70in, left=1in, right=1in]{geometry}
\renewcommand{\baselinestretch}{1.1}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{multicol}
\usepackage{tcolorbox}
\usepackage{caption}
\usepackage{todonotes}
\usepackage[font={sf,footnotesize}]{caption}
\usepackage{hyperref}
\newcommand{\llabel}[1]{\hypertarget{lintarget:#1}{}\linelabel{lin:#1}}

@@

\llabel{inferentialgoal}At first glance, these two modeling paradigms may appear to pursue independent inferential goals--- explanation vs.\ prediction---but in reality this distinction is blurred. Trend estimation may be nominally focused on explanatory outcomes, but especially because it tends to be carried out using simpler models, it often lacks a clear link with theories that facilitate meaningful causal interpretation. Moreover, whether implicitly or explicitly, estimated trends are frequently treated as quantities that project forward into the future, taking on a forecasting role. Meanwhile, although more complex forecasting models are certainly built for predictive purposes, the typical use of mechanistic, process-based model components is often justified by the assumption that ``we must understand to predict”---and indeed, such models are often judged by how faithfully they mimic the real-world processes they represent---which contributes to the conflation of explanation with prediction \citep{Shmueli2010}.

@@

\begin{figure}
	\centering
	\includegraphics{../figures/figure_modeldataspaces_revised}
	\caption{Different trajectories through model-data space during model development, from pre-model/pre-data to post-model/post-data quadrants.
	Panel A serves as a legend, showing where model development should begin (carefully thinking about the hypotheses we want to test) and when inference should be made, for either explanatory purpose (estimation) or forecasting (prediction).
	Panel B illustrates that in current trend estimation approaches, most time is spent in the post-model/post-data quadrant, where models are directly fitted to empirical data---often with no prior evaluation of the model.
	Panel C illustrates the typical process-based model development. A lot of time is spent building and calibrating (fitting) separate submodels (represented by the separated ovals), without clear distinction between model building and model calibration and without evaluating the full model structure a priori. Additionally, the specific hypotheses the model is designed to test are rarely clearly stated.
	Panel D illustrates the workflow highlighted in this paper (see main text and Figure 2 for details). Ideally, it begins in the pre-model/pre-data quadrant, where the research question is defined. The next step moves to the post-model/pre-data quadrant, where a model is built---independently of the data. This is where we should spend a good deal of time evaluating model behavior using simulated data. Once we are satisfied with our model, we can move to the post-model/post-data quadrant, where the model is fitted to real data. Then, we perform retrodictive checks to compare model predictions to observations, which likely give us some feedback to refine our model.}
	\label{fig:modeldata}
\end{figure}

@@

@@

\begin{figure}[h]
	\centering
	\vspace*{-0.5cm}
    \hspace*{-1.5cm}
	\includegraphics{../figures/figure_worflow_wsteps}
	\caption{The workflow we propose here \citep[which builds from recent advances in workflows,][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024} focuses on iterative feedbacks between the research question, model and data. Most research should start at Step 1, with the research question, followed by extensive model building and evaluation using simulated data (Steps 2 and 3) before proceeding to fitting the model on empirical data and examining it through retrodictive checks (Steps 4 and 5). Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.  The integration of observations (the data stream) occurs at Step 4---only after the model has been thoroughly evaluated---and can also highlight opportunities to collect new data and enrich the data stream. This figure is expanded from Figure \ref{fig:modeldata}d.}
	\vspace*{-0cm}
	\label{fig:workflow}
\end{figure}

@@

@@

Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to accurately \llabel{quant1}understand and manage uncertainties and yield new scientific insights. This will help refocus the debate on designing new hypotheses, formulating new questions---and guiding efforts to collect new data. Here, we illustrate how such a workflow could lead to significant improvements in two case studies: (i) estimating global biodiversity trends and (ii) forecasting future species and ecosystem dynamics using process-based models.

@@

In principle, increasing model complexity can be beneficial by enhancing mechanistic realism, better representing more complex dynamics, and \llabel{quant2}reducing uncertainty by restricting the effective parameter space. In practice, however, this is not always the outcome. In process-based modeling for climate forecasts, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong disagreements exist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017}.
These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
Some researchers now advocate for simplifying models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. 
If a model becomes too complex, it may become a black box,
and understanding the sources of uncertainty and how they propagate through the model may become nearly impossible (see Box).
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.

@@

Applying the workflow to process-based models would help open the black box.
Each successive step of model development in the workflow may highlight current problems and a path to solutions. Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two, and help expose potential parameter identifiability issues in the model design. 
Uncovering identifiability issues would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity. The workflow may also highlight model problems by requiring more explicit model calibration (i.e. data fitting), which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model. Non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity). The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating it through projections.

@@

% Trends - outline current problem
% In the current workflow for estimating trends over time, a new model with a new estimate often leads to a paper (see figure A above) because ecologists spend less time interrogating their models with simulated data, or their model performance fit to empirical data.
In the current workflow for estimating trends over time, a model is typically chosen a priori to fit to data, yielding parameter estimates that are treated as true and often translated directly into published findings and conclusions, without interrogating the model using simulated data and carefully evaluating its performance. Any such analysis must be treated with caution, because different model variants can yield very different estimates (hence different conclusions), and published results may not prove robust when confronted with future reanalysis of the same data.
The Living Planet Index (LPI, \url{www.livingplanetindex.org}), which aims to include long-term data on vertebrate populations of species across the globe, is emblematic of these conflicting results. % \todo[]{could highlight how influential the LPI has become (recent reports suggest a 73\% ave. pop decline since 1970. This would be massive!}
% With updated data released semi-annually alongside new datasets and new estimates of decline, 
A growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach. For examples using LPI data, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed many previous results \citep{Johnson2024}.

@@

% We suggest unified training in our proposed workflow, or a similar one \citep{betanworkflow,Gelman2020, grinsztajn2021,vandeschoot2021}, would focus on learning to generate questions and then models, and then how to simulate data from them. 
\llabel{collab}Although in some cases it may be most expedient for ecologists to collaborate with statisticians who are already well-equipped to adopt our proposed workflow or a similar workflow \citep{betanworkflow,Gelman2020, grinsztajn2021,vandeschoot2021}, we recommend that domain scientists who regularly engage in model-driven research efforts receive unified training about how to generate well-formed questions and appropriate models, and then how to simulate data from them.
Through this and the use retrodictive checks, most ecologists would be better equipped to think through what parameters are most critical to their question and/or aim (e.g., management), and also gain a much stronger connection to the level of uncertainty in many of ecological estimates. Empiricists would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models. Process-based modelers may start a new generation of simpler models that are more tractable to theoretical ecologists, who may see new bridges from their work to empirical data and forecasting.

Beyond routine application of the workflow and universal training to support it, the full transformative potential of the workflow hinges on additional broader systemic changes within the research community. First, fostering more open model sharing through standardized repositories and collaborative platforms would accelerate scientific discovery by allowing researchers to build upon, validate, and refine existing models more efficiently, Second, open sharing of more comprehensive study design and sampling design details associated with datasets would enable more appropriate data integration and model design decisions, and independent validation of these decisions by the community. Third, where feasible, alignment on common datasets would provide a standardized basis for model development and comparison, reducing the fragmentation of efforts and facilitating more direct and meaningful evaluations of different modeling approaches. Coupled with universal training, these shifts in the sharing and standardization of data and models would create a more collaborative, efficient, and rigorous environment for ecological modeling, ultimately leading to more accurate forecasts and better understanding of complex environmental systems.

% New paragraph by Jim!
\llabel{studydesignandmore}Beyond routine application of the workflow and universal training to support it, the full transformative potential of the workflow hinges on additional broader systemic changes within the research community. First, fostering more open model sharing through standardized repositories and collaborative platforms would accelerate scientific discovery by allowing researchers to build upon, validate, and refine existing models more efficiently, Second, open sharing of more comprehensive study design and sampling design details associated with datasets would enable more appropriate data integration and model design decisions, and independent validation of these decisions by the community. Third, where feasible, alignment on common datasets would provide a standardized basis for model development and comparison, reducing the fragmentation of efforts and facilitating more direct and meaningful evaluations of different modeling approaches. Coupled with universal training, these shifts in the sharing and standardization of data and models would create a more collaborative, efficient, and rigorous environment for ecological modeling, ultimately leading to more accurate forecasts and better understanding of complex environmental systems.

\subsection{[Need a better title] A tractable alternative to machine learning}
